---
title: "Corona Virus Sim"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      message=FALSE,
                      warning=FALSE)

library(tidyverse)
library(scales)
library(lubridate)

options(scipen = 999)

```

## Attempt to simulation potential trajectories for Corona Virus.

Consider factors that will affect spread of virus:

- number of cases at any given time
- daily % increase in cases, and how that changes over time
- R0: rate at which virus spreads; on average, how many others are infected by each person who gets the disease

### Simple, limited example

#### Set up parameters

```{r}
## set parameters
#ncases <- 100 added in loop
rate_d_m <- 0.2
rate_d_sd <- 0.08

## number of days to project
days <- 30
```


```{r}
df_vir_growth <- data.frame()
i <- 1
## create loop for single simulation
  ncases <- 100
  df_vir_growth_sim <- data.frame(sim=i, day=0, cases=ncases, newcases=100)
  df_vir_growth <- bind_rows(df_vir_growth, df_vir_growth_sim)
  last_cases <- ncases
for(d in 1:days){
  #last <- nrow(df_vir_growth)
  #last_cases <- df_vir_growth[last,3]
  new_cases <- last_cases * rnorm(mean=rate_d_m, sd=rate_d_sd, n=1) 
  ttl_cases <- last_cases+new_cases
  df_vir_grow <- data.frame(sim=i, day=d, cases=ttl_cases+new_cases, newcases=new_cases)
  df_vir_growth <- bind_rows(df_vir_growth, df_vir_grow)
  last_cases <- ttl_cases
}
df_vir_growth$sim <- as.factor(df_vir_growth$sim)  
```


```{r}
#ggplot(df_vir_growth, aes(x=day, y=cases, color=sim))+geom_line()
```

### Create simulation with multiple examples

```{r}
## originally copied from prev chunk

nsims <- 20 ## number of simulations
#s <- 1 ## starting sim identifier number

## set up empty data frame with needed cols
#df_vir_growth <- data.frame(sim=1, day=0, cases=ncases, newcases=100)
df_vir_growth <- data.frame()

## create loop for each sim
for(i in 1:nsims){
## create loop for single simulation
  ncases <- 100
  df_vir_growth_sim <- data.frame(sim=i, day=0, cases=ncases, newcases=100, rate=1)
  df_vir_growth <- bind_rows(df_vir_growth, df_vir_growth_sim)
  last_cases <- ncases
for(d in 1:days){
  #last <- nrow(df_vir_growth)
  #last_cases <- df_vir_growth[last,3]
  grate <- max(0.01, rnorm(mean=rate_d_m, sd=rate_d_sd, n=1))
  new_cases <- last_cases * grate
  ttl_cases <- last_cases+new_cases
  df_vir_grow <- data.frame(sim=i, day=d, cases=ttl_cases+new_cases, newcases=new_cases, rate=grate)
  df_vir_growth <- bind_rows(df_vir_growth, df_vir_grow)
  last_cases <- ttl_cases
}

}
#df_vir_growth$day <- as.factor(df_vir_growth$day)
df_vir_growth$sim <- as.factor(df_vir_growth$sim)
```

```{r}
## get means for each week
df_vir_smry_day <- df_vir_growth %>%  
  group_by(day) %>% summarize(mean_cases=mean(cases),
                              sd_cases=sd(cases),
                              iqlow_cases=quantile(cases, 0.25),
                              iqup_cases=quantile(cases, 0.75))
```


```{r}

ggplot(df_vir_growth, aes(x=day, y=cases, color=sim))+geom_line()+
  geom_line(df_vir_smry_day, mapping=aes(x=day, y=mean_cases, color='mean'), size=2)+
  scale_y_continuous(labels=comma)

```

#### Summary stats at final day in sim:

```{r}
df_vir_smry <- df_vir_growth %>% filter(day==days) %>% 
  group_by(day) %>% summarize(mean_cases=mean(cases),
                              sd_cases=sd(cases),
                              iqlow_cases=quantile(cases, 0.25),
                              iqup_cases=quantile(cases, 0.75))
```

At day `r df_vir_smry$day`: <br />

Mean cases: `r df_vir_smry$mean_cases` <br />
IQR: `r df_vir_smry$iqlow_cases` - `r df_vir_smry$iqup_cases` <br />
1 sd below: `r df_vir_smry$mean_cases - df_vir_smry$sd_cases` <br />
1 sd above: `r df_vir_smry$mean_cases + df_vir_smry$sd_cases`

Boxplot view to show distribution by day:

```{r}
df_vir_growth$day <- as.factor(df_vir_growth$day)
ggplot(df_vir_growth, aes(x=day, y=cases))+geom_boxplot()
```

