---
title: "Distribution Exploration"
output:
  html_document:
    df_print: paged
---

## Exploring distributions and related probability concepts {.tabset}

```{r INITIAL, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      message=FALSE,
                      warning=FALSE)

library(tidyverse)
library(lubridate)
library(scales)

```

### Normal

Key concepts, using the Normal distribution as a starting point.

#### Generate Normal Distribution

Generate a random set of numbers that follow a specified normal distribution.

```{r GENERATE NORMAL}

m <- 50
std <- 10
norm_01 <- rnorm(n=1000, mean=m, sd=std)
```

Normal distribution with:

* Mean: `r m` 
* Std. deviation: `r std`

```{r VIEW NORMAL}

plot(norm_01)
hist(norm_01)
boxplot(norm_01)

```

Convert to data frame use with ggplot and additional features therein. Get summary statistics.

```{r NORM DATA FRAME}
df_norm_01 <- data.frame(values=norm_01)

summary(df_norm_01)
print(paste0('Std dev: ', sd(df_norm_01$values)))
```

```{r NORM GGPLOT 1}

ggplot(df_norm_01, aes(x=values))+geom_histogram(fill='grey')+
  geom_vline(xintercept = mean(df_norm_01$values), color='green', linetype='dashed', size=1)+
  geom_vline(xintercept = median(df_norm_01$values), color='blue', linetype='dotted')+
  geom_vline(xintercept=quantile(df_norm_01$values, 0.25), color='purple')+
  geom_vline(xintercept = quantile(df_norm_01$values, 0.75), color='purple')+
  geom_vline(xintercept=mean(df_norm_01$values)-sd(df_norm_01$values), color='orange')+
  geom_vline(xintercept=mean(df_norm_01$values)+sd(df_norm_01$values), color='orange')+
  scale_y_continuous(expand=c(0,0))+
  scale_x_continuous(expand=c(0,0))+
  theme_classic()
```

* Green = mean
* Blue = median
* Purple = 25th and 75th percentiles
* Orange = 1 std dev, above and below mean

Convert histogram to density function, with probabilities on y-axis:

```{r}
df_norm_01 <- df_norm_01 %>% mutate(
  prob=dnorm(values, mean=m, sd=std)
)

ggplot(df_norm_01, aes(x=values, y=prob))+geom_line()+
  geom_vline(xintercept = mean(df_norm_01$values), color='green', linetype='dashed', size=1)+
  geom_vline(xintercept = median(df_norm_01$values), color='blue', linetype='dotted')+
  geom_vline(xintercept=quantile(df_norm_01$values, 0.25), color='purple')+
  geom_vline(xintercept = quantile(df_norm_01$values, 0.75), color='purple')+
  geom_vline(xintercept=mean(df_norm_01$values)-sd(df_norm_01$values), color='orange')+
  geom_vline(xintercept=mean(df_norm_01$values)+sd(df_norm_01$values), color='orange')+
  scale_y_continuous(labels=percent)+
  theme_classic()

```

Plot the cumulative distrbution function for this data:

* can be read as cumulative percentiles on the left

```{r}
plot(ecdf(df_norm_01$values))
```

Can create the same by applying pnorm() to the data and drawing with ggplot2:

```{r}
df_norm_01 <- df_norm_01 %>% mutate(
  pctl = pnorm(values, mean=m, sd=std)
)

ggplot(df_norm_01, aes(x=values, y=pctl))+geom_line()+
  geom_vline(xintercept = mean(df_norm_01$values), color='green', linetype='dashed', size=1)+
  geom_vline(xintercept = median(df_norm_01$values), color='blue', linetype='dotted')+
  geom_vline(xintercept=quantile(df_norm_01$values, 0.25), color='purple')+
  geom_vline(xintercept = quantile(df_norm_01$values, 0.75), color='purple')+
  geom_vline(xintercept=mean(df_norm_01$values)-sd(df_norm_01$values), color='orange')+
  geom_vline(xintercept=mean(df_norm_01$values)+sd(df_norm_01$values), color='orange')+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(labels=percent, expand=c(0,0), limits=c(0,1))+
  theme_classic()

```

#### Probabilities

What is the probability of value X as an outcome?

```{r PROB 1}
options(scipen = 99)
x <- 40
dnorm(x=x, mean=m, sd=std)

```

A single number in a continous distribution, so the probability will be low.

What is the probability of an outcome of X or lower?

```{r PROB 2}
x <- mean(df_norm_01$values)-sd(df_norm_01$values)
pnorm(q=x, mean=m, sd=std)

```

what is the probability of an outcome between X and Y?

```{r PROB 3}
x <- x
y <- mean(df_norm_01$values)+sd(df_norm_01$values)
p_lower <- pnorm(q=x, mean=m, sd=std)
p_upper <- pnorm(q=y, mean=m, sd=std, lower.tail=FALSE)
pxy <- 1-(p_lower+p_upper)

```

Probability of an outcome between `r x` and `r y` is: `r pxy`.

### Poisson

* for discrete random variable to determine probability of x occurences of an event over specific interval of time
* works best with small number of events (10 or less) in small time interval (1 or less)
* at smaller valuse of lambda (<10), has strong right skew
* at larger values of lambda, approaches normal - probably because at smaller values, there is more potential upside than downside, since can't be negative 

#### Generate Poisson

```{r}
nn <- 10000
lmb <- 5
pois_01 <- rpois(n=nn, lambda=lmb)
hist(pois_01)
```

#### Probability distribution

```{r}
xx <- c(1:max(pois_01))
lmb <- 5
pois_02 <- dpois(x=xx, lambda=lmb)
plot(pois_02)
```

#### Example

https://www.youtube.com/watch?v=m_c1nIC2Dfo 

* website gets average 10 visits per minute

```{r}
nn <- 1000
lmb <- 10
wpois <- rpois(n=nn, lambda = lmb)
hist(wpois)

xx <- c(1:20)
wpois <- dpois(x=xx, lambda = lmb)
plot(wpois)
```

* what is probability of 3 visits?
* what is probability of 3 visits or less?
* what is probability of more than 10 visits?

```{r}
w1 <- 3
lmb <- 10
wdpois <- dpois(x=w1, lambda=lmb)
print(paste0("3 visits: ", wdpois*100, " percent"))

wppois <- ppois(q=w1, lambda = lmb)
print(paste0(w1, " visits or less: ", wppois*100, " percent"))

w1 <- 12
wppois <- ppois(q=w1, lambda = lmb, lower.tail=FALSE)
print(paste0("More than ",w1, " visits: ", wppois*100, " percent"))

```

Based on the above distribution, what if you wanted to know how many visits per day?

* create data frame for distribution based on minute
* multiply up
* (not sure if this is legit, though?)

```{r}
vpm <- 10
df_pois <- data.frame(visits=c(1:24))
#df_pois$prob <- dpois(x=df_pois$visits, lambda=10)
df_pois <- df_pois %>% mutate(
  prob=dpois(x=visits, lambda=vpm),
  cprob=ppois(q=visits, lambda=vpm)
)

plot(df_pois$prob)
plot(df_pois$cprob)

df_pois <- df_pois %>% mutate(
  visitsd=visits*60*24
)

ggplot(df_pois, aes(x=visitsd, y=prob))+geom_line()
```

Or to simulate the data...

```{r}
nn <- 1000
df_pois_sim <- data.frame(vpms=rpois(n=nn, lambda=vpm))
df_pois_sim <- df_pois_sim %>% mutate(
  vpd=vpms*60*24
)
ggplot(df_pois_sim, aes(x=vpms))+geom_histogram(binwidth = 1)
## set bindwidth by calc
bw <- (max(df_pois_sim$vpd)-min(df_pois_sim$vpd))/15
## plot vpd
ggplot(df_pois_sim, aes(x=vpd))+geom_histogram(binwidth=bw)
```


### Gamma family

Gamma is actually a family of distributions. It seems to offer better replication of skewed distributions than Poisson.

From <https://www.rocscience.com/help/swedge/swedge/Gamma_Distribution.htm>:

* The Gamma distribution is widely used in engineering, science, and business, to model continuous variables that are always positive and have skewed distributions.
* a (alpha) is known as the shape parameter, while b (beta) is referred to as the scale parameter.
* b has the effect of stretching or compressing the range of the Gamma distribution. A Gamma distribution with b = 1 is known as the standard Gamma distribution.

The Gamma distribution represents a family of shapes. As suggested by its name, a controls the shape of the family of distributions. The fundamental shapes are characterized by the following values of a:

Case I (a < 1)

* When a < 1, the Gamma distribution is exponentially shaped and asymptotic to both the vertical and horizontal axes.

Case II (a = 1)

* Gamma distribution with shape parameter a = 1 and scale parameter b is the same as an exponential distribution of scale parameter (or mean) b.

Case III (a > 1)

* When a is greater than one, Gamma distribution assumes a mounded (unimodal), skewed shape. Skewness reduces as the value of a increases.

There are two parameters and two variations on the parameter combinations. For simplicity, focus here is on:

* shape (a above, sometimes k) -> determines skew
* scale (b above, sometimes theta) -> determines stretch/compression of the range (can also use rate which equals 1/scale)

Some gamma distribution examples:

**(what does it mean when probability density >1 ???)**

1. Do not confuse 'probability density' with probability
2. Calculus
3. In practical terms, it can be interpreted as something like "at really small intervals, there is a high amount of density (frequency) here, *relative* to the overall distribution of values, and as a result some values have frequency of (close to) 0." 

```{r GAMMA DIST}

## function to create gamma samples and density
fGammavardens <- function(num, sh, sc){
  shsc <- paste0('gamma','-',sh,':', sc)
  cat(paste0('shape=',sh,' rate=', sc))
  datagam <- rgamma(n=1000, shape=sh, scale=sc)
  plot.hist <- hist(datagam)
  #plot.hist <- ggplot(data.frame(datagam), aes(x=datagam))+geom_histogram()
  
  datagam <- round(datagam, 3)
  datagam <- unique(datagam)
  dengam <- dgamma(datagam, shape=sh, scale=sc)
  gam <- data.frame(datagam, dengam, ksc=shsc)
  plot.dense <- plot(data=gam, dengam~datagam)
  #plot.dense <- ggplot(gam, aes(x=datagam, y=dengam))+geom_line()+theme_classic()
  
  list(plot.hist, plot.dense)
  return(gam)
}

gam_all <- data.frame()

gamsample <- fGammavardens(1000, 1, 0.5)
gam_all <- bind_rows(gam_all, gamsample)

gamsample <- fGammavardens(1000, 1, 1)
gam_all <- bind_rows(gam_all, gamsample)

gamsample <- fGammavardens(1000, 1, 2)
gam_all <- bind_rows(gam_all, gamsample)

gamsample <- fGammavardens(1000, 2, 0.5)
gam_all <- bind_rows(gam_all, gamsample)

gamsample <- fGammavardens(1000, 2, 2)
gam_all <- bind_rows(gam_all, gamsample)

gamsample <- fGammavardens(1000, 3, 0.5)
gam_all <- bind_rows(gam_all, gamsample)

gamsample <- fGammavardens(1000, 3, 1)
gam_all <- bind_rows(gam_all, gamsample)

gamsample <- fGammavardens(1000, 5, 1)
gam_all <- bind_rows(gam_all, gamsample)


gam_all <- gam_all %>% distinct(datagam, .keep_all = TRUE)
ggplot(gam_all, aes(x=datagam, y=dengam, color=ksc))+
  geom_line()

```

### Exponential

As noted above, this is part of gamma. Treated here separately for additional exploration.

Inspired by info here:
https://web.stanford.edu/class/bios221/labs/simulation/Lab_3_simulation.html 

Exponential distribution functions:

* dexp, pexp, qexp, rexp

Because it is specific type of distribution (part of the gamma family), it has specific parameters (I guess). For rexp:

* 'n' appears to be equivalent to 'shape'; looks like '1' might be min value
* 'rate' corresponds to 1/scale in gamma distribution

```{r}
reps <- 5000
nexps <- 1 ## 1=full exponential; 2 highly right skewed; 20 almost normal
rate <- 0.1
set.seed(2019)
x1 <- replicate(n=reps, sum(rexp(n=nexps, rate=rate)))
```

Visualize results

```{r, echo=TRUE}
ggplot(data.frame(x1), aes(x=x1))+
  geom_histogram(aes(y=..density..))+theme_classic()

ggplot(data.frame(x1), aes(x=x1))+
  geom_density()+theme_classic()

```

With comparison to theorectical distribution

```{r, echo=TRUE}
ggplot(data.frame(x1), aes(x=x1))+
  geom_histogram(aes(y=..density..))+
  stat_function(fun=function(x)dgamma(x, shape=nexps, scale=1/rate), color='red', size=1)+
  theme_classic()
```

### Beta

Beta has two required parameters and one optional:

* shape1
* shape2
* ncp: non-centrality (optional)

Seems as though:

* shape1 < 1 < shape2: right skew
* shape1 > 1 > shape2: left skew equivalent to right skew on same values swapped
* strength of skew generally reflective of ratio of higher shape to lower
* shape1, shape2 = 1: pretty much uniform
* shape1, shape2 = 3+: approaches normality


```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 0.25
sh2 <- 4
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density(color='yellow', size=1)
```

```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 0.25
sh2 <- 1
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density()
```

```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 0.5
sh2 <- 1
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density()
```

```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 0.5
sh2 <- 5
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density()
```

```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 1
sh2 <- 1
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density()
```

```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 1
sh2 <- 2
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density()
```

```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 2
sh2 <- 1
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density()
```

```{r, echo=TRUE, fig.height=4, fig.width=4}
num_sample <- 1000
sh1 <- 5
sh2 <- 5
ggplot(data.frame(rb=rbeta(n=num_sample, shape1=sh1, shape2 = sh2)), aes(x=rb))+geom_histogram(aes(y=..density..))+
  geom_density()
```

Duplicate beta distribution example for a/b test from <https://www.countbayesie.com/blog/2015/3/3/6-amazing-trick-with-monte-carlo-simulations>

For distribution of each test:

* shape1 = # of conversions
* shape2 = # not converted 
* peak of curve is 

```{r, echo=TRUE, fig.height=4, fig.width=6}
num_sample <- 50000

# a test
convert <- 20
notconvert <- 100
acr <- convert/sum(convert, notconvert)

a <- rbeta(n=num_sample, shape1=convert, shape2=notconvert)
ad <- density(a)
#plot(ad)

# b test
convert <- 38
notconvert <- 110
bcr <- convert/sum(convert, notconvert)
b <- rbeta(n=num_sample, shape1=convert, shape2=notconvert)
bd <- density(b)
#plot(bd)

abd <- data.frame(ax=ad$x,
                  ay=ad$y,
                  bx=bd$x,
                  by=bd$y)

ggplot(abd)+geom_line(aes(x=ax, y=ay), color='blue')+
  geom_line(aes(x=bx, y=by), color='red')+
  scale_x_continuous(limits=c(0,1))+
  theme_classic()

## alternate approach to organizing data for display (tidy)
ad <- data.frame(test='a',
                  x=ad$x,
                  y=ad$y
                  )
bd <- data.frame(test='b',
                 x=bd$x,
                 y=bd$y)
abd <- bind_rows(ad, bd)

ggplot(abd, aes(x=x, y=y, color=test))+geom_line()+
  geom_vline(xintercept=acr, linetype='dashed')+
  geom_vline(xintercept=bcr, linetype='dashed')+
  scale_x_continuous(limits=c(0,1))+
  theme_classic()
```

Stats from a/b test:

* A conversion rate: `r acr`
* B conversion rate: `r bcr`

Conversion rates very close to peak probability in chart (x-axis).