---
title: 'Sim Experiment: weekly forecast'
output:
  html_document:
    df_print: paged
---

```{r include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      message=FALSE,
                      warning=FALSE)

library(tidyverse)
library(scales)

options(scipen=999)

```

## Intro

Experimenting with using simulation to create a week-by-week forecast, based on historical data.

### Sampling Method

Use sampling of probabilities based on historical data to create a forecast for benchmarks.

### 1. Generate some example historical data

#### basic sample data

* for each metric, get numbers for each week picked randomly from normal distribution with set mean and sd 

```{r EXAMPLE}
set.seed(2019)

## generate weeks
wk_start <- -12
wk <- seq(from=wk_start, to=0, by=1)

## generate sample with increase
m1 <- 1000
sd1 <- 200
c01 <- rnorm(n=length(wk), mean=m1, sd=sd1)

m2 <- 1200
sd2 <- 300
c02 <- rnorm(n=length(wk), mean=m2, sd=sd2)

comparison <- data.frame(week=wk,
                         comp01=c01,
                         comp02=c02)

ggplot(comparison, aes(x=wk, y=comp01))+geom_line(aes(color='comp01'))+
  geom_line(y=comparison$comp02, aes(color='comp02'))+
  scale_y_continuous(labels=comma, limits=c(0,max(comparison$comp02)), expand=c(0,0))+
  scale_x_continuous(labels=comma, expand=c(0,0) )+
  theme_classic()

```

Results of data sampling for normal distribution:

* comp01: mean=`r m1`, sd=`r sd1`
* comp02: mean=`r m2`, sd=`r sd2`

#### adjust for weeks
 
* Activity starts slow and ramps up to week 0
* Model this by multiplying value for each week by 1/abs(w), where 'w' is week number (absolute value needed because weeks are negative)
* Since last week is 0, ratio calc will give infinity (Inf) so replace with 2x existing value

```{r}
## divide 
comparison <- comparison %>% mutate(
  comp01a=comp01*(1/abs(wk)),
  comp02a=comp02*(1/abs(wk))
)

comparison <- comparison %>% mutate(
  comp01a = ifelse(comp01a==Inf, comp01*2, comp01a),
  comp02a = ifelse(comp02a==Inf, comp02*2, comp02a)
)

ggplot(comparison, aes(x=wk, y=comp01a))+geom_line(aes(color='comp01a'))+
  geom_line(y=comparison$comp02a, aes(color='comp02a'))+
  scale_y_continuous(labels=comma, limits=c(0,max(comparison$comp02a)), expand=c(0,0))+
  scale_x_continuous(labels=comma, expand=c(0,0) )+
  theme_classic()

```

### 2. Use historical example to create new sim data

```{r}
## gather comparison data to facilitate sample comparison
comp_orig <- comparison %>% select(week, comp01a, comp02a) %>% gather(
  key=sample, value=amount, -week
)
num_sample <- 500
samp_data_all <- data.frame()

for(s in 1:num_sample){
#  for each row in comparison table, get a sample of data points
#  randomly chosen from between comp01a and comp02a with leeway on either side
  leewaydown <- 0.9
  leewayup <- 1.1
  for(w in 1:length(wk)){
    dist <- c(comparison$comp01a[w],comparison$comp02a[w])
    samp_data <- data.frame(week=wk[w], 
                            sample=paste0('s',s), 
                            amount=runif(n=1, min=min(dist)*leewaydown, max=max(dist)*leewayup),
                            stringsAsFactors = FALSE)
    samp_data_all <- bind_rows(samp_data_all, samp_data) 
  }
}

```

* for each week, get a randomly sampled value based on uniform dist between the existing historical values
* with some leeway allowed: `r leewaydown` of lower value, `r leewayup` of upper value
* number of samples: `r num_sample`

Combine sim data with original comparisons

```{r}
comp_all <- bind_rows(comp_orig, samp_data_all)
```

Visualize results

```{r}
## line chart with original comparisons highlighted in larger size
ggplot(data=comp_orig, aes(x=week, y=amount, color=sample))+geom_line(size=1.2, linetype='dashed')+
   geom_line(data=comp_all, aes(x=week, y=amount, color=sample))+
  scale_y_continuous(labels=comma, limits=c(0,max(comparison$comp02a)), expand=c(0,0))+
  scale_x_continuous(labels=comma, expand=c(0,0) )+
  theme_classic()+theme(legend.position = 'none')
```

### 3. Analyze totals, means and distributions

```{r}
comp_summary <- comp_all %>% group_by(sample) %>% summarize(
  ttl=sum(amount),
  mean=mean(amount),
  median=median(amount),
  stddev=sd(amount)
)
```

```{r}
ggplot(comp_summary, aes(x=ttl))+geom_histogram()+
  geom_vline(aes(xintercept=comp_summary$ttl[comp_summary$sample=='comp01a']), linetype='dotted', color='blue')+
  geom_vline(aes(xintercept=comp_summary$ttl[comp_summary$sample=='comp02a']), linetype='dotted', color='blue')+
  geom_vline(aes(xintercept=mean(comp_summary$ttl)), linetype='dashed', color='green')+
  geom_vline(aes(xintercept=mean(comp_summary$ttl)-(2*sd(comp_summary$ttl))), linetype='dashed', color='red')+
  geom_vline(aes(xintercept=mean(comp_summary$ttl)+(2*sd(comp_summary$ttl))), linetype='dashed', color='red')+
  theme_classic()
```

(green line is average) <br />
(blue lines represent totals for original comparison data sets)<br />
(red lines represent 2 sd above/below mean)

```{r}
summary(comp_summary$ttl)

msd2plus <- mean(comp_summary$ttl)+(2*sd(comp_summary$ttl))
msd2min <- mean(comp_summary$ttl)-(2*sd(comp_summary$ttl))

msd05 <- qnorm(p=.025, mean=mean(comp_summary$ttl), sd=sd(comp_summary$ttl))
msd95 <- qnorm(p=.025, mean=mean(comp_summary$ttl), sd=sd(comp_summary$ttl), lower.tail=FALSE)

```

Based on sample distribution:

* 95% chance total will be between: `r msd2min` and `r msd2plus`

Theoretically, based on normal distribution with same mean and sd:

* 95% chance between `r msd05` and `r msd95`

```{r, echo=TRUE}
## what amount is at pth percentile?
qnorm(p=0.5, mean=mean(comp_summary$ttl), sd=sd(comp_summary$ttl))
quantile(comp_summary$ttl, 0.5)
```

```{r, echo=TRUE}

## probability of exactly x
dnorm(x=mean(comp_summary$ttl), mean=mean(comp_summary$ttl), sd=sd(comp_summary$ttl))
## probability of q or less
pnorm(q=mean(comp_summary$ttl), mean=mean(comp_summary$ttl), sd=sd(comp_summary$ttl))
```

### Smoothing Method

Similar to what we are currently using: build a model based on smoothing of historical references

### 1. Apply smoothed line to chart

* need to remove the 'color' parameter that differentiates the two comparison sets in order to create one smooth line (otherwise, will show smooth line for each comparison set)

```{r}
ggplot(comp_orig, aes(x=week, y=amount))+geom_point()+
  geom_smooth(method='loess')+
  scale_y_continuous(labels=comma, expand=c(0,0))+
  scale_x_continuous(labels=comma, expand=c(0,0) )+
  theme_classic()

```

### 2. Create predictive model based on smoothing

```{r}

model_sm <- loess(amount~week, data=comp_orig)
summary(model_sm)

predict_sm <- predict(model_sm, data.frame(week=wk))

predict_sm_df <- data.frame(week=wk,
                            sample='model',
                            amount=predict_sm, stringsAsFactors = FALSE)

## linear model for comparison
# model_lm <- lm(amount~week, data=comp_orig)
# summary(model_lm)
# predict(model_lm, data.frame(week=-12), interval='confidence')
# predict(model_lm, data.frame(week=-12), interval='prediction')
# pred_lm <- predict(model_lm, data.frame(week=wk), interval='confidence')

```

Add model predictions to new data frame, with original historical data

```{r}
comp_model <- bind_rows(comp_orig, predict_sm_df)

ggplot(comp_model, aes(x=week, y=amount, color=sample))+geom_line()+
  geom_line(data=predict_sm_df, aes(x=week, y=amount), size=1.2)+
  theme_classic()
  
```

### 3. Check total amount from model

* Compare to simulation mean of `r mean(comp_summary$ttl)`

```{r}
sum(predict_sm_df$amount)

```

### 4. Get confidence intervals for loess prediction

(doable but somewhat tricky...for another time)
